---
title: "Multivariate linear regression"
author: "Joaquín Martínez-Minaya"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multivariate linear regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

This vignette demonstrates how to use the **multivarbayes** package to fit a Bayesian multivariate linear regression model, summarize the results, plot the posterior distributions, and make predictions on new data.

### Example Usage



```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(multivarbayes)
library(ggplot2) # for plotting
library(dplyr)
library(tidyr)
```

### Generate Example Data

We generate some example data with an intercept term and multivariate normal errors.

```{r, warning=FALSE, message = FALSE}
set.seed(123)
n <- 1000  # number of observations
k <- 3    # number of covariates (including intercept)
m <- 2    # number of response variables
nsims <- 1000

# Covariate matrix with intercept
X <- cbind(1, matrix(rnorm(n * (k - 1)), n, k - 1))  

# True coefficients
B_true <- matrix(c(1, 0.5, -0.3, 2, -0.5, 1), ncol = m)

# Multivariate normal errors
Sigma_true <- matrix(c(1, 0.3, 0.3, 1), ncol = m)
errors <- MASS::mvrnorm(n, mu = rep(0, m), Sigma = Sigma_true)

# Response matrix
Y <- X %*% B_true + errors  

# Combine into data frame
data <- data.frame(cbind(Y, X))
colnames(data) <- c("Y1", "Y2", "Intercept", "X1", "X2")
```

### Fit the Model

We fit the model using the `mlvr` function and the formula interface.

```{r}
formula <- as.matrix(data[, c("Y1", "Y2")]) ~  X1 + X2
model_fit <- mlvr(formula, data = data)
model_fit$analytic
```

### Summarize the Fitted Model

We can summarize the fitted model to inspect the posterior distributions of the coefficients and the scale matrix of Sigma.

```{r}
summary(model_fit)
```

### Plot Posterior Distributions

To visualize the posterior distributions of the coefficients, we can use the `plot` method.

```{r}
plot(model_fit)

```


```{r}
summary(model_fit)
B_true
```


### Predict New Observations

Finally, we can use the `predict` function to predict responses for new observations based on the fitted model.

```{r}

# Example usage:
# Assuming model_fit is the fitted object and newdata is the new data to predict on
newdata <- data.frame(X1 = c(0.5, -1, 2), X2 = c(-0.5, 1.5, 0))  # New covariate values
newdata <- data[1:5, c("X1", "X2")]
newdata_real <- data[1:5, c("Y1", "Y2")]

predictions <- predict(model_fit, newdata = newdata)
#cat("--- Y1 --- \n")
predictions$summary$Y1 %>% round(.,3)

#cat("--- Y2 --- \n")
predictions$summary$Y2 %>% round(., 3)
```
In this final part, we compare the theoretical predictive distribution with the simulated predictive distribution, as well as with the actual observed values. This allows us to visually inspect how well the model captures the underlying data distribution and the uncertainty associated with it.


```{r, warning = FALSE, message=FALSE}
# Suppose `predictions` is the object containing the results obtained from the `predict.mlvr` function
# and `newdata_real` holds the actual values for each predicted individual.
# Extract necessary elements from predictions
analytic_mean <- predictions$analytic$mean          # Theoretical predictive mean
analytic_cov <- predictions$analytic$cov_matrix     # Theoretical predictive covariance
predictive_samples <- predictions$marginals         # Simulated predictive samples
df <- predictions$analytic$df                       # Degrees of freedom for the t-Student distribution
newdata_real <- newdata_real                        # Actual values provided

# Create a data frame to store simulated predictive samples
simulated_data <- data.frame()
n_obs <- nrow(analytic_mean)                        # Number of observations in newdata
n_sims <- ncol(predictive_samples[[1]])             # Number of simulations (assume all responses have the same number of simulations)
n_responses <- length(predictive_samples)           # Number of response variables

# Build the data frame for simulated predictive samples
for (resp in 1:n_responses) {
  for (obs in 1:n_obs) {
    sim_values <- predictive_samples[[resp]][obs, ]  # Simulated samples for the response variable and observation
    sim_df <- data.frame(
      Observation = paste0("Obs_", obs),             # Observation label
      Response = paste0("Y", resp),                  # Response variable name
      Value = sim_values,                            # Simulated samples
      Group = "Simulated"                            # Group label
    )
    simulated_data <- rbind(simulated_data, sim_df)
  }
}

# Create a data frame for theoretical densities
theoretical_data <- data.frame()

# Calculate theoretical density for each observation and response variable
for (obs in 1:n_obs) {
  for (resp in 1:n_responses) {
    mean_val <- analytic_mean[obs, resp]          # Theoretical mean for the response variable
    var_val <- analytic_cov[resp, resp, obs]      # Variance for the response variable
    
    # Create a sequence of values to represent the theoretical density
    x_vals <- seq(mean_val - 4 * sqrt(var_val), mean_val + 4 * sqrt(var_val), length.out = 500)
    
    # Calculate density using the t-Student distribution
    dens_vals <- dt((x_vals - mean_val) / sqrt(var_val), df = df) / sqrt(var_val)
    
    # Create a data frame to store theoretical densities
    theor_df <- data.frame(
      Observation = paste0("Obs_", obs),
      Response = paste0("Y", resp),
      Value = x_vals,
      Density = dens_vals,
      Group = "Theoretical"
    )
    theoretical_data <- rbind(theoretical_data, theor_df)
  }
}

# Create a data frame for the actual observed values
real_values_data <- data.frame(
  Observation = rep(paste0("Obs_", 1:n_obs), each = n_responses),  # Labeled observations
  Response = rep(paste0("Y", 1:n_responses), times = n_obs),       # Response variable name
  Value = as.vector(t(newdata_real)),                              # Convert actual values to a vector
  Group = "Real"                                                   # Group label for "Real"
)

# Convert simulated data to densities for plotting
simulated_densities <- simulated_data %>%
  group_by(Observation, Response) %>%
  do({
    density_data <- density(.$Value, adjust = 1.2)  # Calculate density for simulated samples
    data.frame(Value = density_data$x, Density = density_data$y)  # Store density values
  }) %>%
  ungroup() %>%
  mutate(Group = "Simulated")  # Label as "Simulated" for plotting

# Combine theoretical and simulated densities into a single data frame
plot_data <- rbind(simulated_densities, theoretical_data)

# Create plots
ggplot() +
  # Histograms for simulated predictive samples
  geom_line(data = simulated_densities, aes(x = Value, y = Density, color = Group), size = 1.2) +
  # Line for theoretical densities
  geom_line(data = theoretical_data, aes(x = Value, y = Density, color = Group), linetype = "dashed", size = 1.2) +
  # Vertical line for actual observed values
  geom_vline(data = real_values_data, aes(xintercept = Value, color = Group), linetype = "dotted", size = 1.2) +
  facet_grid(Response ~ Observation, scales = "free") +  # Facets by response and observation
  labs(title = "Predictive Distributions for Each Observation",
       x = "Value",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "top", legend.title = element_text(face = "bold")) +
  scale_color_manual(values = c("blue", "red", "green")) +   # Colors for distributions and actual values
  guides(color = guide_legend(title = "Distribution Types"))

```



